\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-03-20 analysis/paper/paper.tex
still working on poptime section

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras; the interested
reader can refer to Saarela \& Arjas \citeyearpar{saarela2015non} and
Saarela \citeyearpar{saarela2016case}. First, let
\(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the
event of interest for individual \(i=1, \ldots,n\). For simplicity, we
will consider Type I censoring due to the end of follow-up at time
\(\tau\) (the general case of non-informative censoring is treated in
Saarela \citeyearpar{saarela2016case}). We assume a continuous time
model, which implies that the counting process jumps are less than or
equal to one. We are interested in modeling the hazard functions
\(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)].\]

Next, we model the base series sampling mechanism using non-homogeneous
Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\), with the
person-moments where \(dR_i(t) = 1\) constituting the base series. The
process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and
base series person-moments contributed by individual \(i\). This process
is typically defined by the user via its hazard function \(\rho_i(t)\).
The process \(Q_{i}(t)\) is characterized by
\(E[dQ_{i}(t)] = \lambda_{i}(t)dt + \rho_i(t)dt\).

If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
terms of \(\theta\), we could define an estimator \(\hat{\theta}\) by
maximization of the likelihood expression
\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},\]
where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
\(u\). However, the integral over time makes the computation and
maximisation of \(L_0(\theta)\) challenging.

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual contributions of the form
\[P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
Therefore, we can define an estimating equation for \(\theta\) as
follows:
\[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
When a logarithmic link function is used for modeling the hazard
function, the above expression is of a logistic regression form with an
offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected
in the case-base sampling mechanism are person-moments, rather than
individuals, and the parameters to be estimated are hazards or hazard
ratios rather than odds or odds ratios. Generally, an individual can
contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the logarithm of this
estimating function has mean zero at the true value \(\theta=\theta_0\),
and that the resulting estimator \(\hat{\theta}\) is asymptotically
normally distributed.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest sampling the base series \emph{uniformly} from the study base.
In terms of Poisson processes, this sampling strategy corresponds
essentially to a time-homogeneous Poisson process with hazard equal to
\(\rho_i(t) = b/B\), where \(b\) is the number of sampled observations
in the base series, and \(B\) is the total population-time for the study
base. More complex examples are also available; see for example Saarela
\& Arjas \citeyearpar{saarela2015non}, where the probabilities of the
sampling mechanism are proportional to the cardiovascular disease event
rate given by the Framingham score. With this sampling mechanism,
Saarela \& Arjas are able to increase the efficiency of their
estimators, when compared to uniform sampling.

Let \(g(t; X)\) be the linear predictor such that
\(\log(\lambda(t;X)) = g(t; X)\). Different functions of \(t\) lead to
different parametric hazard models. The simplest of these models is the
one-parameter exponential distribution which is obtained by taking the
hazard function to be constant over the range of \(t\):

\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 X.\]

@@

\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.\]

@@

\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \]

@@

\[\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]

where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(\lambda_j(t)\) is the corresponding hazard function. As
may be expected, this functional form is similar to the terms appearing
in the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with some of the
functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation for ridge.
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
are the only packages to offer this functionality.
\textcolor{red}{I can dive deeper to determine if this is a fair statement}

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Several \proglang{R} packages implement methodologies for competing risk
analysis; for a different perspective, see Mahani \& Sharabiani
\citeyearpar{mahani2015bayesian}. The package \pkg{cmprsk} provides
methods for cause-specific subdistributions, such as in the Fine-Gray
model \citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific cumulative incidence functions from
unadjusted, non-parametric survival functions. Our package
\pkg{casebase} also provides functionalities for competing risk analysis
by estimating the cause-specific hazards, and from these quantities, we
can derive the cause-specific cumulative incidence functions.

@@

@@

The sampling scheme selected for \code{sampleCaseBase} was described in
Hanley \& Miettinen \citeyearpar{hanley2009fitting}: we first sample
along the ``person'' axis, proportional to each individual's total
follow-up time, and then we sample a moment uniformly over their
follow-up time. This sampling scheme is equivalent to the following
picture: imagine representing the total follow-up time of all
individuals in the study along a single dimension, where the follow-up
time of the next individual would start exactly when the follow-up time
of the previous individual ends. Then the base series could be sampled
uniformly from this one-dimensional representation of the overall
follow-up time. In any case, the output is a dataset of the same class
as the input, where each row corresponds to a person-moment. The
covariate profile for each such person-moment is retained, and an offset
term is added to the dataset. This output could then be used to fit a
smooth hazard function, or for visualization of the base series.

The fitting function \code{fitSmoothHazard} starts by looking at the
class of the dataset: if it was generated from \code{sampleCaseBase}, it
automatically inherited the class \code{cbData}. If the dataset supplied
to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the
fitting function starts by calling \code{sampleCaseBase} to generate the
base series. In other words, users can bypass \code{sampleCaseBase}
altogether and only worry about the fitting function
\code{fitSmoothHazard}.

The fitting function retains the familiar formula interface of
\code{glm}. The left-hand side of the formula should be the name of the
column corresponding to the event type. The right-hand side can be any
combination of the covariates, along with an explicit functional form
for the time variable. Note that non-proportional hazard models can be
achieved at this stage by adding an interaction term involving time
(cf.~Case Study 4 below). The offset term does not need to be specified
by the user, as it is automatically added to the formula before calling
\code{glm}.

@@

\[ S(t) = \exp\left(-\int_0^t \Lambda(u;X) du\right),\qquad \Lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
From this, we can derive the event-specific subdensities:

\[ f_j(t) = \lambda_j(t)S(t).\]

@@

\[ CI_j(t) = \int_0^t f_j(u)du.\] The integrals are computed using
either numerical integration (via the trapezoidal rule) or Monte Carlo
integration. This option is controlled by the argument \code{method} of
the \code{absoluteRisk} function.

We created \code{absoluteRisk} as an \code{S3} generic, with methods
that match the different types of outputs of \code{fitSmoothHazard}. The
method dispatch system of \proglang{R} then takes care of matching the
correct output to the correct methodology for calculating the cumulative
incidence function, without requiring any intervention from the user.

In the following sections, we illustrate these functionalities in the
context of four case studies.

@@

For the first case study, we will use of the European Randomized Study
of Prostate Cancer Screening data. This dataset is available through the
\pkg{casebase} package:

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm,
R>   levels = c(0, 1),
R>   labels = c("Control group", "Screening group")
R> )
\end{CodeInput}
\end{CodeChunk}

The results of this study were published by
\citep{schroder2009screening}, and the dataset itself was obtained using
the approach described in \citep{liu2014recovering}.

Population time plots can be extremely informative graphical displays of
survival data. They should be the first step in an exploratory data
analysis. We facilitate this task in the \pkg{casebase} package using
the \code{popTime} function. We first create the necessary dataset for
producing the population time plots, and we can generate the plot by
using the corresponding \code{plot} method:

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
R> plot(pt_object)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC,
R>   event = "DeadOfPrCa",
R>   exposure = "ScrArm"
R> )
R> plot(pt_object_strat)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
R>   data = ERSPC
R> )
R> summary(cox_model)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(
R>   ScrArm = c("Control group", "Screening group"),
R>   ignore = 99
R> )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>   xlab = "Years since Randomization",
R>   ylab = "Cumulative Incidence (%)",
R>   fun = "event",
R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R>   main = sprintf(
R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>     exp(coef(cox_model)),
R>     exp(confint(cox_model))[1],
R>     exp(confint(cox_model))[2]
R>   ),
R>   ylim = c(0, 2), yscale = 100
R> )
R> legend("topleft",
R>   legend = c("Control group", "Screening group"),
R>   col = c("red", "blue"),
R>   lty = c(1, 1),
R>   bg = "gray90"
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
R>   data = ERSPC,
R>   ratio = 100
R> )
R> 
R> summary(casebase_exponential)
R> exp(coef(casebase_exponential)[2])
R> exp(confint(casebase_exponential)[2, ])
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(
R>   object = casebase_exponential,
R>   time = seq(0, 15, 0.1),
R>   newdata = new_data
R> )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>   xlab = "Years since Randomization",
R>   ylab = "Cumulative Incidence (%)",
R>   fun = "event",
R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R>   main = sprintf(
R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>     exp(coef(cox_model)),
R>     exp(confint(cox_model))[1],
R>     exp(confint(cox_model))[2]
R>   ),
R>   ylim = c(0, 2), yscale = 100
R> )
R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft",
R>   legend = c(
R>     "Control group (Cox)", "Control group (Casebase)",
R>     "Screening group (Cox)", "Screening group (Casebase)"
R>   ),
R>   col = c("red", "red", "blue", "blue"),
R>   lty = c(1, 2, 1, 2),
R>   bg = "gray90"
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
R>   data = ERSPC,
R>   ratio = 100
R> )
R> 
R> summary(casebase_time)
R> exp(coef(casebase_time))
R> exp(confint(casebase_time))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(
R>   object = casebase_time,
R>   time = seq(0, 15, 0.1),
R>   newdata = new_data
R> )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>   xlab = "Years since Randomization",
R>   ylab = "Cumulative Incidence (%)",
R>   fun = "event",
R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R>   main = sprintf(
R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>     exp(coef(cox_model)),
R>     exp(confint(cox_model))[1],
R>     exp(confint(cox_model))[2]
R>   ),
R>   ylim = c(0, 2), yscale = 100
R> )
R> lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
R> lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
R> 
R> legend("topleft",
R>   legend = c(
R>     "Control group (Cox)", "Control group (Casebase)",
R>     "Screening group (Cox)", "Screening group (Casebase)"
R>   ),
R>   col = c("red", "red", "blue", "blue"),
R>   lty = c(1, 2, 1, 2),
R>   bg = "gray90"
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
R>   data = ERSPC,
R>   ratio = 100
R> )
R> 
R> summary(casebase_splines)
R> exp(coef(casebase_splines))
R> exp(confint(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(
R>   object = casebase_splines,
R>   time = seq(0, 15, 0.1),
R>   newdata = new_data
R> )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>   xlab = "Years since Randomization",
R>   ylab = "Cumulative Incidence (%)",
R>   fun = "event",
R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R>   main = sprintf(
R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>     exp(coef(cox_model)),
R>     exp(confint(cox_model))[1],
R>     exp(confint(cox_model))[2]
R>   ),
R>   ylim = c(0, 2), yscale = 100
R> )
R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
R> 
R> legend("topleft",
R>   legend = c(
R>     "Control group (Cox)", "Control group (Casebase)",
R>     "Screening group (Cox)", "Screening group (Casebase)"
R>   ),
R>   col = c("red", "red", "blue", "blue"),
R>   lty = c(1, 2, 1, 2),
R>   bg = "gray90"
R> )
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeInput}
R> anova(casebase_exponential, casebase_splines, test = "LRT")
\end{CodeInput}
\end{CodeChunk}

As expected, we see that splines model provides a better fit. Similarly,
we can compare the AIC for all three case-base models:

\begin{CodeInput}
R> c(
R>   "Exp." = AIC(casebase_exponential),
R>   "Weib." = AIC(casebase_time),
R>   "Splines" = AIC(casebase_splines)
R> )
\end{CodeInput}
\end{CodeChunk}

Once again, we have evidence that the splines model provides the best
fit.

With this first case study, we saw how population-time plots can provide
a useful summary of the incidence for our event of interests. In
particular, we saw how it can provide evidence against an exponential
hazard model. We then explored how different parametric models can
provide different estimates of the CIF, even though they all give
similar estimates for the hazard ratios.

\hypertarget{case-study-2bone-marrow-transplant}{%
\section{Case study 2--Bone-marrow
transplant}\label{case-study-2bone-marrow-transplant}}

In the next case study, we show how case-base sampling can be used in
the context of a competing risk analysis. For illustrative purposes, we
will use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is also available as part of the
\pkg{casebase} package.

\begin{CodeInput}
R> 
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

The data contains information on 177 patients who received a stem-cell
transplant for acute leukemia. The event of interest is relapse, but
other competing causes (e.g.~transplant-related death) were also
recorded. Several covariates were captured at baseline: sex, disease
type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
CR3), source of stem cells (bone marrow and peripheral blood, coded as
BM+PB, or only peripheral blood, coded as PB), and age. A summary of
these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
We note that the statistical summaries were generated differently for
different variable types: for continuous variables, we gave the range,
followed by the mean and standard deviation; for categorical variables,
we gave the counts for each category.

First, we can look at a population-time plot to visualize the incidence
density of both relapse and the competing events. In Figure
\ref{fig:compPop}, failure times are highlighted on the plot using red
dots for the event of interest (panel A) and blue dots for competing
events (panel B). In both panels, we see evidence of a non-constant
hazard function: the density of points is larger at the beginning of
follow-up than at the end.

Our main objective is to compute the absolute risk of relapse for a
given set of covariates. We start by fitting a smooth hazard to the data
using a linear term for time:

>>>>

As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
expanded in Saarela \citeyearpar{saarela2016case}), writing the
likelihood arising from this data-generating mechanism using the
framework of non-homogeneous Poisson processes, we eventually reach an
expression where each person-moment's contribution is of the form
\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
process associated with the event of interest, \(h(t)\) is the
corresponding hazard function, and \(\rho(t)\) is the hazard function
for the Poisson process associated with case-base sampling. This
parametric form suggests that we can readily estimate log-hazards of the
form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
observation corresponds to a person moment, the function \(g(t; X)\) is
linear in a finite number of parameters, and where we treat
\(-\log(\rho(t))\) as an offset.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest sampling the base series \emph{uniformly} from the study base.
In terms of Poisson processes, this sampling strategy corresponds
essentially to a time-homogeneous Poisson process with intensity equal
to \(b/B\), where \(b\) is the number of sampled observations in the
base series, and \(B\) is the total population-time for the study base.
More complex examples are also available; see for example Saarela \&
Arjas \citeyearpar{saarela2015non}, where the probabilities of the
sampling mechanism are proportional to the cardiovascular disease event
rate given by the Framingham score. With this sampling mechanism,
Saarela \& Arjas are able to increase the efficiency of their
estimators, when compared to uniform sampling.

Let \(g(t; X)\) be the linear predictor such that
\(\log(h(t;X)) = g(t; X)\). Different functions of \(t\) lead to
different parametric hazard models. The simplest of these models is the
one-parameter exponential distribution which is obtained by taking the
hazard function to be constant over the range of \(t\):

\[ \log(h(t; X)) = \beta_0 + \beta_1 X.\]

@@

\[ \log(h(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.\]

@@

\[ \log(h(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \]

@@

\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]

where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
be expected, this functional form is similar to the terms appearing in
the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with a description
of some of the functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: \pkg{CoxRidge}
\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
\pkg{survival} \citeyearpar{survival-package}. On the other hand,
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, this is the only package to offer
this functionality.
\textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks. The differentiating factor between these packages and
\pkg{casebase} is that \pkg{casebase} handles competing risks through
multiple logistic regression.
\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}. Could you incorporate some of their discussion, with proper citations? The biggest omission right now is \pkg{cmprsk}, which we actually use below.}

@@

The case-base sampling approach described in Section \ref{theory} can be
visualized in the form of a population time plot. They are extremely
informative graphical displays of survival data and should be one of the
first steps in an exploratory data analysis. We facilitate this task in
the \pkg{casebase} package using the \code{popTime} function and
\code{plot} method as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The \code{casebase::popTime} function takes as input the original
  dataset along with the column names corresponding to the timescale,
  the event status and an exposure group of interest (optional). This
  will create an object of class \code{popTime} (or
  \code{popTimeExposure} if you specify a value for the exposure
  argument).\\
\item
  The corresponding \code{plot} methods for the object created in Step 1
  can be called to create the population time plot with several options
  for customizing the aesthetics.
\end{enumerate}

To illustrate these functions we will use the European Randomized Study
of Prostate Cancer Screening data which is available through the
\pkg{casebase} package. The results of this study were published by
\citep{schroder2009screening}, and the dataset itself was obtained using
the approach described in \citep{liu2014recovering}.

We first create the necessary dataset for producing the population time
plots using the \code{casebase::popTime} function. In this example we
stratify by the treatment group. Note that the resulting object stores
the exposure variable as an attribute:

\begin{CodeOutput}
#> [1] "ScrArm"
\end{CodeOutput}
\end{CodeChunk}

We then pass this object to the \code{plot} function which leverages the
\pkg{ggplot2} package \citep{ggplot2} to build population time plots. It
builds the plot by sequentially adding layers, starting with a layer for
the area representing the population time followed by points to show the
casebase sampling mechanism. This function gives user the flexibility to
add any combination of the case.series, base.series and competing
events. The case series and competing events are sampled at random
vertically on the plot in order to visualise the incidence density using
the popTime function. That is, imagine we draw a vertical line at a
specific event time. We then plot the point at a randomly sampled
y-coordinate along this vertical line. This is done to avoid having all
points along the upper edge of the plot (because the subjects with the
least amount of observation time are plotted at the top of the y-axis).
By randomly distributing them, we can get a better sense of the
inicidence density. The base series is sampled horizontally on the plot
using the sampleCaseBase function.

One benefit from these plots is that it allows you to see the incidence
density. This can be seen from the distribution of the red dots in the
above plot. We can see that more events are observed later on in time.
Therefore a constant hazard model would not be appropriate in this
instance as it would overestimate the cumulative incidence earlier on in
time, and underestimate it later on.

The unique \emph{step shape} of the population time plot is due to the
randomization of the Finnish cohorts which were carried out on January 1
of each of the 4 years 1996 to 1999. This, coupled with the uniform
December 31 2006 censoring date, lead to large numbers of men with
exactly 11, 10, 9 or 8 years of follow-up.

It is important to note that the red points are random distributed
across the grey area for each time of event. That is, imagine we draw a
vertical line at a specific event time. We then plot the red point at a
randomly sampled y-coordinate along this vertical line. This is done to
avoid having all red points along the upper edge of the plot (because
the subjects with the least amount of observation time are plotted at
the top of the y-axis). By randomly distributing them, we can get a
better sense of the inicidence density.

\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}

For a more detailed description of how to modify the aesthetics of a
population time plot, please refer to the
\href{http://sahirbhatnagar.com/casebase/articles/customizingpopTime.html}{Customizing
Population Time Plots} vignette.

@@

The sampling scheme selected for \code{sampleCaseBase} was described in
Hanley and Miettinen \citeyearpar{hanley2009fitting}: we first sample
along the ``person'' axis, proportional to each individual's total
follow-up time, and then we sample a moment uniformly over their
follow-up time. This sampling scheme is equivalent to the following
picture: imagine representing the total follow-up time of all
individuals in the study along a single dimension, where the follow-up
time of the next individual would start exactly when the follow-up time
of the previous individual ends. Then the base series could be sampled
uniformly from this one-dimensional representation of the overall
follow-up time. In any case, the output is a dataset of the same class
as the input, where each row corresponds to a person-moment. The
covariate profile for each such person-moment is retained, and an offset
term is added to the dataset. This output could then be used to fit a
smooth hazard function, or for visualization of the base series.

The fitting function \code{fitSmoothHazard} starts by looking at the
class of the dataset: if it was generated from \code{sampleCaseBase}, it
automatically inherited the class \code{cbData}. If the dataset supplied
to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the
fitting function starts by calling \code{sampleCaseBase} to generate the
base series. In other words, the occasional user can bypass
\code{sampleCaseBase} altogether and only worry about the fitting
function \code{fitSmoothHazard}.

The fitting function retains the familiar formula interface of
\code{glm}. The left-hand side of the formula should be the name of the
column corresponding to the event type. The right-hand side can be any
combination of the covariates, along with an explicit functional form
for the time variable. Note that non-proportional hazard models can be
achieved at this stage by adding an interaction term involving time. The
offset term does not need to be specified by the user, as it is
automatically added to the formula.

@@

\[ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).\]
From this, we can derive the event-specific subdensities:

\[ f_j(t) = h_j(t)S(t).\]

@@

\[ CI_j(t) = \int_0^t f_j(u)du.\]

We created \code{absoluteRisk} as an \code{S3} generic, with methods for
the different types of outputs of \code{fitSmoothHazard}. The method
dispatch system of \proglang{R} then takes care of matching the correct
output to the correct methodology for calculating the cumulative
incidence function, without the user's intervention.

In the following sections, we illustrate these functionalities in the
context of three case studies.

@@

@@

\begin{CodeOutput}
#> Call:
#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
#>     ScrArm, data = ERSPC)
#> 
#>   n= 159893, number of events= 540 
#> 
#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#>                       exp(coef) exp(-coef) lower .95 upper .95
#> ScrArmScreening group     0.801       1.25     0.674     0.952
#> 
#> Concordance= 0.519  (se = 0.011 )
#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
#> Wald test            = 6.37  on 1 df,   p=0.01
#> Score (logrank) test = 6.39  on 1 df,   p=0.01
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{center}\includegraphics{../figures/unnamed-chunk-1-1} \end{center}

\hypertarget{equivalence-between-casebase-and-aft}{%
\subsection{Equivalence between casebase and
AFT}\label{equivalence-between-casebase-and-aft}}

@@

@@

@@

@@

@@

@@

@@

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:unnamed-chunk-4}Mean }
\centering
\begin{tabular}[t]{>{\bfseries}lcc}
\toprule
Model & casebase & survreg\\
\midrule
Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
\cmidrule{1-3}
Gompertz & 0.79 (0.66, 0.94) & 0.80 (0.67, 0.95)\\
\cmidrule{1-3}
Weibull & 0.80 (0.67, 0.95) & 0.80 (0.65, 0.96)\\
\cmidrule{1-3}
Splines & 0.80 (0.67, 0.95) & --\\
\bottomrule
\multicolumn{3}{l}{\textit{Note:}}\\
\multicolumn{3}{l}{Median (Inter-quartile range) is given for Model Size.}\\
\end{tabular}
\end{table}

@@
